---
title: "THRIVE Analysis Final Draft"
author: "Amia Graye"
output: html_document
date: "2026-03-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

```{r libraries}
# load the libraries
library(readr)
library(readxl)
library(dplyr)
library(lme4)
library(lmerTest)
library(ggplot2)
library(tidyr)
library(mice)
library(geepack)
library(stringr)
library(emmeans)
```

```{r data}
# load the data
dash_drs <- read_csv("Desktop/JHU HW/Consulting_Data/Data/DASH Rapid Screener (DRS).csv")
bp_meas <- read_csv("Desktop/JHU HW/Consulting_Data/Data/Blood Pressure Measurement Form.csv")
thrive_random <- read_excel("Desktop/JHU HW/Consulting_Data/Data/THRIVE Randomization Data.xlsx")
lab_results <- read_csv("Desktop/JHU HW/Consulting_Data/Data/Laboratory Results_updated.csv")
elig <- read.csv("~/Desktop/JHU HW/Consulting_Data/Data/Study Team Eligibility Form.csv")
med_history <- read.csv("~/Desktop/JHU HW/Consulting_Data/Data/Medical History Follow-up (12W,.csv")
fim04 <- read.delim("~/Desktop/JHU HW/Consulting_Data/Data/FIMrecalls04.txt")
demo <- read_csv("Desktop/JHU HW/Consulting_Data/Data/Sociodemographic Questionnaire .csv")
bp_drugs <- read_excel("Desktop/JHU HW/Consulting_Data/Data/nhanes_antihypertensives_2017_2020_ss_edit_RO.xlsx")

# change the data locations for your specific computer
```

# Cleaning

## Blood Pressure Measurements

```{r bp meas cleaning}
# clean the blood pressure data
df <- bp_meas %>%
  select(Case, bp_event, bp_avgsbp, bp_avgdbp) %>%
  rename(caselabel = Case) %>%
  left_join(thrive_random, by = "caselabel") %>%
  mutate(
    bp_event = factor(bp_event,
                      levels = c("Baseline", "12W", "24W"))
  ) %>%
  filter(!is.na(Randomization))
```

## Blood Pressure Medications

```{r bp med cleaning}
# clean the blood pressure medication data
# change names of dataset to match Grace's code
baseline <- elig
followup <- med_history

# change caselabel column to Case in randomization data (already loaded)
THRIVE_Randomization_Data <- thrive_random %>%
  rename(Case = caselabel) 

# remove unnecessary columns from baseline and followup data
baseline <- baseline %>%
  select(Case, eligibilitybl_exclusionmedications, eligibilitybl_exclusionmedsdm, eligibilitybl_exclusionmedsdmtype, eligibilitybl_exclusionmedsktype, eligibilitybl_exclusionmedsktype_note) 

followup <- followup %>%
  select(Case, medhistfu_event, medhistfu_htnmeds, medhistfu_hbpmeds, medhistfu_hbpmedcount, medhistfu_hbpmedname) %>%
  rename(bp_event = medhistfu_event)

# add randomization data to baseline and followup data
baseline <- baseline %>%
  left_join(THRIVE_Randomization_Data, by = "Case") %>%
  rename(caselabel = Case)
followup <- followup %>%
  left_join(THRIVE_Randomization_Data, by = "Case") %>%
  rename(caselabel = Case)
```

```{r bp med cleaning 2}
# create vector of all meds used for hypertension treatment in the dataset
bp_drug_names <- bp_drugs %>%
  filter(`Used for Hypertension treatment` == "Yes") %>%
  pull(RXDDRUG) %>%
  str_to_upper()

# convert meds in baseline and followup to uppercase
baseline <- baseline %>%
  mutate(
    med_name = str_to_upper(eligibilitybl_exclusionmedsktype_note)
  )

followup <- followup %>%
  mutate(
    med_name = str_to_upper(medhistfu_hbpmedname)
  )

# create baseline bp indicator variable
baseline <- baseline %>%
  mutate(
    bp_med_baseline =
      ifelse(
        eligibilitybl_exclusionmedsktype == "Yes" &
        str_detect(med_name, str_c(bp_drug_names, collapse = "|")),
        1, 0
      )
  )

# create followup bp indicator variable
followup <- followup %>%
  mutate(
    bp_med_followup =
      ifelse(
        (medhistfu_htnmeds == "Yes" | medhistfu_hbpmeds == "Yes") &
        str_detect(med_name, str_c(bp_drug_names, collapse = "|")),
        1, 0
      )
  )

# add indictor for baseline visit
baseline <- baseline %>%
  mutate(bp_event = "Baseline")

# combine the baseline and followup datasets
medication_visits <- bind_rows(
  baseline %>% select(caselabel, bp_event, bp_med_baseline) %>%
    rename(bp_med = bp_med_baseline),

  followup %>% select(caselabel, bp_event, bp_med_followup) %>%
    rename(bp_med = bp_med_followup)
)

# add randomization data
medication_visits <- medication_visits %>%
  mutate(
    bp_event = factor(bp_event,
                      levels = c("Baseline", "12W", "24W"))
  ) %>%
  left_join(thrive_random %>% select(caselabel, Randomization), by = "caselabel") %>%
  filter(!is.na(Randomization)) %>%
  arrange(caselabel, bp_event)

# manually change some values for participants according to Shanshan's notes
## 18398-125 - no 12W med
medication_visits <- medication_visits %>%
  mutate(
    bp_med = ifelse(caselabel == "18398-125" & bp_event == "12W", 0, bp_med)
  )

## 18398-363 - no 24W med
medication_visits <- medication_visits %>%
  mutate(
    bp_med = ifelse(caselabel == "18398-363" & bp_event == "24W", 0, bp_med)
  )
```

```{r bp med cleaning 3}
# combine medication data with main dataset
df <- df %>%
  left_join(medication_visits, by = c("caselabel", "bp_event")) %>%
  mutate(
    bp_med = ifelse(is.na(bp_med), 0, bp_med)
  )

# check if numbers match up with Grace's table
medication_visits %>%
  group_by(bp_event, Randomization) %>%
  summarise(
    n_on_med = sum(bp_med == 1, na.rm = TRUE),
    n_total = n(),
    percent = 100 * n_on_med / n_total
  ) %>% 
  arrange(bp_event)
```

## DASH Adherence Score

```{r dash cleaning}
first_non_missing_chr <- function(x) {
  x <- as.character(x)
  x <- x[!is.na(x) & x != ""]
  if (length(x) == 0) return(NA_character_)
  x[[1]]
}

first_non_missing_num <- function(x) {
  x <- as.numeric(x)
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_real_)
  x[[1]]
}

safe_mean <- function(x) {
  x <- as.numeric(x)
  if (all(is.na(x))) return(NA_real_)
  mean(x, na.rm = TRUE)
}

score_mellen <- function(v, full, mid, direction = c("low", "high")) {
  direction <- match.arg(direction)
  if (direction == "low") {
    ifelse(is.na(v), NA_real_, ifelse(v <= full, 1, ifelse(v <= mid, 0.5, 0)))
  } else {
    ifelse(is.na(v), NA_real_, ifelse(v >= full, 1, ifelse(v >= mid, 0.5, 0)))
  }
}
```

```{r dash cleaning 2}
# Randomized ITT cohort
rand <- thrive_random %>%
  clean_names() %>%
  transmute(
    case = as.character(caselabel),
    arm_raw = as.character(randomization),
    arm = case_when(
      stringr::str_detect(arm_raw, "Intervention") ~ "Intervention",
      TRUE ~ "Control"
    )
  ) %>%
  mutate(arm = factor(arm, levels = c("Control", "Intervention"))) %>%
  select(case, arm) %>%
  distinct()

stopifnot(n_distinct(rand$case) == 80L)

# Baseline covariates from sociodemographic form
Sociodemographic_Questionnaire <- read_csv("Desktop/JHU HW/Consulting_Data/Data/Sociodemographic Questionnaire .csv")
sd <- Sociodemographic_Questionnaire %>%
  clean_names() %>%
  transmute(
    case = as.character(case),
    demo_age,
    demo_sexatbirth,
    demo_ishispaniclatino
  ) %>%
  group_by(case) %>%
  summarise(
    age = first_non_missing_num(demo_age),
    sex = first_non_missing_chr(demo_sexatbirth),
    hispanic = first_non_missing_chr(demo_ishispaniclatino),
    .groups = "drop"
  ) %>%
  mutate(
    sex = factor(if_else(sex %in% c("Female", "Male"), sex, NA_character_), levels = c("Female", "Male")),
    hispanic = factor(if_else(hispanic %in% c("No", "Yes"), hispanic, NA_character_), levels = c("No", "Yes"))
  )

# SBP outcomes
Blood_Pressure_Measurement_Form <- read_csv("Desktop/JHU HW/Consulting_Data/Data/Blood Pressure Measurement Form.csv")
bp_long <- Blood_Pressure_Measurement_Form  %>%
  clean_names() %>%
  transmute(
    case = as.character(case),
    visit = as.character(bp_event),
    sbp = rowMeans(cbind(bp_sbp1, bp_sbp2, bp_sbp3), na.rm = TRUE)
  ) %>%
  mutate(sbp = if_else(is.nan(sbp), NA_real_, sbp)) %>%
  filter(case %in% rand$case, visit %in% c("Baseline", "12W", "24W")) %>%
  group_by(case, visit) %>%
  summarise(sbp = first_non_missing_num(sbp), .groups = "drop")

sbp_wide <- bp_long %>%
  pivot_wider(names_from = visit, values_from = sbp) %>%
  rename(
    sbp_bl = Baseline,
    sbp_12 = `12W`,
    sbp_24 = `24W`
  )

# DASH nutrient score (Mellen): recall-level from FIMrecalls04
visit_map <- c("0" = "Baseline", "12" = "12W", "24" = "24W")
FIMrecalls04 <- read.delim("~/Desktop/JHU HW/Consulting_Data/Data/FIMrecalls04.txt")
fim04 <- FIMrecalls04 %>%
  clean_names() %>%
  transmute(
    case = as.character(participant_id),
    visit_num = as.character(visit_number),
    kcal = as.numeric(energy_kcal),
    caloriesfromsfa = as.numeric(x_calories_from_sfa),
    caloriesfromfat = as.numeric(x_calories_from_fat),
    caloriesfromprotein = as.numeric(x_calories_from_protein),
    cholesterolmg = as.numeric(cholesterol_mg),
    totaldietaryfiberg = as.numeric(total_dietary_fiber_g),
    magnesiummg = as.numeric(magnesium_mg),
    calciummg = as.numeric(calcium_mg),
    potassiummg = as.numeric(potassium_mg),
    sodiummg = as.numeric(sodium_mg)
  ) %>%
  filter(case %in% rand$case, visit_num %in% names(visit_map), !is.na(kcal), kcal > 0)

dash_recall <- fim04 %>%
  mutate(
    chol_per1000kcal = cholesterolmg / kcal * 1000,
    fiber_per1000kcal = totaldietaryfiberg / kcal * 1000,
    magnesium_per1000kcal = magnesiummg / kcal * 1000,
    calcium_per1000kcal = calciummg / kcal * 1000,
    potassium_per1000kcal = potassiummg / kcal * 1000,
    sodium_per1000kcal = sodiummg / kcal * 1000,
    sfa_dash = score_mellen(caloriesfromsfa, full = 6, mid = 11, direction = "low"),
    fat_dash = score_mellen(caloriesfromfat, full = 27, mid = 32, direction = "low"),
    protein_dash = score_mellen(caloriesfromprotein, full = 18, mid = 16.5, direction = "high"),
    cholesterol_dash = score_mellen(chol_per1000kcal, full = 71.4, mid = 107.1, direction = "low"),
    fiber_dash = score_mellen(fiber_per1000kcal, full = 14.8, mid = 9.5, direction = "high"),
    magnesium_dash = score_mellen(magnesium_per1000kcal, full = 238, mid = 158, direction = "high"),
    calcium_dash = score_mellen(calcium_per1000kcal, full = 590, mid = 402, direction = "high"),
    potassium_dash = score_mellen(potassium_per1000kcal, full = 2238, mid = 1534, direction = "high"),
    sodium_dash = score_mellen(sodium_per1000kcal, full = 1143, mid = 1286, direction = "low"),
    dash_nutrient = sfa_dash + fat_dash + protein_dash + cholesterol_dash + fiber_dash +
      magnesium_dash + calcium_dash + potassium_dash + sodium_dash,
    visit = recode(visit_num, !!!visit_map)
  )

dash_visit <- dash_recall %>%
  group_by(case, visit_num, visit) %>%
  summarise(
    dash_score = mean(dash_nutrient, na.rm = TRUE),
    n_recalls_visit = n(),
    dash_category = if_else(dash_score < 4.5, "Low", "High"),
    .groups = "drop"
  )

dash_wide <- dash_visit %>%
  select(case, visit, dash_score) %>%
  pivot_wider(names_from = visit, values_from = dash_score) %>%
  rename(
    dash_bl = Baseline,
    dash_12 = `12W`,
    dash_24 = `24W`
  ) %>%
  left_join(rand %>% select(case, arm), by = "case")
```


# Analysis

## Blood Pressure Measurements

### Evaluate Missingness

Is missingness of average SBP at follow-up visits related to baseline average SBP?

```{r bp meas missingness}
visit_completeness <- df %>%
  mutate(has_sbp = !is.na(bp_avgsbp)) %>%
  group_by(caselabel, bp_event) %>%
  summarise(
    has_visit = any(has_sbp),
    .groups = "drop"
  ) %>%
  tidyr::pivot_wider(
    names_from = bp_event,
    values_from = has_visit,
    values_fill = FALSE
  ) %>%
  mutate(
    visit_pattern = case_when(
      Baseline & `12W` & `24W`              ~ "baseline_12w_24w",
      Baseline & `12W` & !`24W`             ~ "baseline_12w_only",
      Baseline & !`12W` & `24W`             ~ "baseline_24w_only",
      Baseline & !`12W` & !`24W`            ~ "baseline_only",
      TRUE                                  ~ "other"
    )
  )

baseline_only <- df %>%
  filter(bp_event == "Baseline") %>%
  select(caselabel, baseline_sbp = bp_avgsbp) %>%
  left_join(
    visit_completeness %>% 
      select(caselabel, visit_pattern),
    by = "caselabel"
  )
table(baseline_only$visit_pattern)

anova_model <- aov(baseline_sbp ~ visit_pattern, data = baseline_only)
summary(anova_model)

# No statistically significant difference in baseline SBP across visit-completion patterns
# Baseline SBP does not predict who drops out or misses visits
# Strongly supports a Missing At Random (MAR) assumption
```



### Linear Mixed Effects Model

#### SBP: Combined Population

```{r lme sbp combined}
# fit a linear mixed effects model for SBP in the combined population
m_sbp <- lmer(bp_avgsbp ~ bp_event + (1 | caselabel), data = df)
summary(m_sbp)
confint(m_sbp)   # CIs for baseline mean + changes vs baseline
anova(m_sbp)     # overall time effect test
```

#### SBP: By Randomization Group

```{r lme sbp by group}
# fit a linear mixed effects model for SBP by randomization group
m_sbp_group <- lmer(bp_avgsbp ~ bp_event * Randomization + (1 | caselabel), data = df)
summary(m_sbp_group)
confint(m_sbp_group)   
anova(m_sbp_group)     

# fit an estimated marginal means model to get group-specific estimates
emm_sbp <- emmeans(m_sbp_group, ~ Randomization * bp_event)
emm_sbp

change_sbp <- contrast(
  emm_sbp,
  method = "trt.vs.ctrl",
  ref = "Baseline",
  by = "Randomization"
)

summary(change_sbp, infer = TRUE)

# fit a difference-in-differences contrast to compare changes in SBP between groups
did_sbp <- contrast(
  emm_sbp,
  interaction = "revpairwise"
)

summary(did_sbp, infer = TRUE)
```

#### SBP: By Medication Use

```{r lme sbp by med}
# fit a linear mixed effects model for SBP by medication use
m_sbp_med <- lmer(bp_avgsbp ~ bp_event * bp_med + (1 | caselabel), data = df)
summary(m_sbp_med)
confint(m_sbp_med)
anova(m_sbp_med)

# fit an estimated marginal means model to get medication-specific estimates
emm_sbp_med <- emmeans(m_sbp_med, ~ bp_med * bp_event)
emm_sbp_med

change_sbp_med <- contrast(
  emm_sbp_med,
  method = "trt.vs.ctrl",
  ref = "Baseline",
  by = "bp_med"
)

summary(change_sbp_med, infer = TRUE)

# fit a difference-in-differences contrast to compare changes in SBP between medication groups
did_sbp_med <- contrast(
  emm_sbp_med,
  interaction = "revpairwise"
)

summary(did_sbp_med, infer = TRUE)
```

#### SBP: By Randomization Group and Medication Use

```{r lme sbp by group and med}
# fit a linear mixed effects model for SBP by randomization group and medication use
m_sbp_group_med <- lmer(bp_avgsbp ~ bp_event * Randomization + bp_med + (1 | caselabel), data = df)
summary(m_sbp_group_med)
confint(m_sbp_group_med)
anova(m_sbp_group_med)

# fit an estimated marginal means model to get group estimates with medication adjustment
emm_sbp_group_med <- emmeans(m_sbp_group_med, ~ Randomization * bp_event)
emm_sbp_group_med

change_sbp_group_med <- contrast(
  emm_sbp_group_med,
  method = "trt.vs.ctrl",
  ref = "Baseline",
  by = "Randomization"
)

summary(change_sbp_group_med, infer = TRUE)

# fit a difference-in-differences contrast to compare changes in SBP between groups with medication adjustment
did_sbp_group_med <- contrast(
  emm_sbp_group_med,
  interaction = "revpairwise"
)

summary(did_sbp_group_med, infer = TRUE)
```

#### DBP: Combined Population

```{r lme dbp combined}
# fit a linear mixed effects model for DBP in the combined population
m_dbp <- lmer(bp_avgdbp ~ bp_event + (1 | caselabel), data = df)
summary(m_dbp)
confint(m_dbp)   # CIs for baseline mean + changes vs baseline
anova(m_dbp)     # overall time effect test
```

#### DBP: By Randomization Group

```{r lme dbp by group}
# fit a linear mixed effects model for DBP by randomization group
m_dbp_group <- lmer(bp_avgdbp ~ bp_event * Randomization + (1 | caselabel), data = df)
summary(m_dbp_group)
confint(m_dbp_group)
anova(m_dbp_group)

# fit an estimated marginal means model to get group-specific estimates
emm_dbp <- emmeans(m_dbp_group, ~ Randomization * bp_event)
emm_dbp

change_dbp <- contrast(
  emm_dbp,
  method = "trt.vs.ctrl",
  ref = "Baseline",
  by = "Randomization"
)

summary(change_dbp, infer = TRUE)

# fit a difference-in-differences contrast to compare changes in DBP between groups
did_dbp <- contrast(
  emm_dbp,
  interaction = "revpairwise"
)

summary(did_dbp, infer = TRUE)
```

#### DBP: By Medication Use

```{r lme dbp by med}
# fit a linear mixed effects model for DBP by medication use
m_dbp_med <- lmer(bp_avgdbp ~ bp_event * bp_med + (1 | caselabel), data = df)
summary(m_dbp_med)
confint(m_dbp_med)
anova(m_dbp_med)

# fit an estimated marginal means model to get medication-specific estimates
emm_dbp_med <- emmeans(m_dbp_med, ~ bp_med * bp_event)
emm_dbp_med

change_dbp_med <- contrast(
  emm_dbp_med,
  method = "trt.vs.ctrl",
  ref = "Baseline",
  by = "bp_med"
)

summary(change_dbp_med, infer = TRUE)

# fit a difference-in-differences contrast to compare changes in DBP between medication groups
did_dbp_med <- contrast(
  emm_dbp_med,
  interaction = "revpairwise"
)

summary(did_dbp_med, infer = TRUE)
```

#### DBP: By Randomization Group and Medication Use

```{r lme dbp by group and med}
# fit a linear mixed effects model for DBP by randomization group and medication use
m_dbp_group_med <- lmer(bp_avgdbp ~ bp_event * Randomization + bp_med + (1 | caselabel), data = df)
summary(m_dbp_group_med)
confint(m_dbp_group_med)
anova(m_dbp_group_med)

# fit an estimated marginal means model to get group estimates with medication adjustment
emm_dbp_group_med <- emmeans(m_dbp_group_med, ~ Randomization * bp_event)
emm_dbp_group_med

change_dbp_group_med <- contrast(
  emm_dbp_group_med,
  method = "trt.vs.ctrl",
  ref = "Baseline",
  by = "Randomization"
)

summary(change_dbp_group_med, infer = TRUE)

# fit a difference-in-differences contrast to compare changes in DBP between groups with medication adjustment
did_dbp_group_med <- contrast(
  emm_dbp_group_med,
  interaction = "revpairwise"
)

summary(did_dbp_group_med, infer = TRUE)
```

### Data Visualizations

```{r bp plots}
# sbp plot
df %>%
  group_by(bp_event, Randomization) %>%
  summarise(
    mean_sbp = mean(bp_avgsbp, na.rm = TRUE),
    se_sbp   = sd(bp_avgsbp, na.rm = TRUE) / sqrt(n())
  ) %>%
  ggplot(aes(bp_event, mean_sbp, color = Randomization, group = Randomization)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_sbp - se_sbp,
                    ymax = mean_sbp + se_sbp),
                width = 0.1) +
  labs(
    title = "Mean Systolic Blood Pressure Over Time by Study Arm",
    x = "Visit",
    y = "Mean Systolic Blood Pressure (mmHg)",
    color = "Study Arm"
  ) +
  theme_minimal()

sbp_12w <- df %>%
  filter(bp_event == "12W") %>%
  group_by(Randomization) %>%
  summarise(mean_sbp_12w = mean(bp_avgsbp, na.rm = TRUE))

sbp_24w <- df %>%
  filter(bp_event == "24W") %>%
  group_by(Randomization) %>%
  summarise(mean_sbp_24w = mean(bp_avgsbp, na.rm = TRUE))

# dbp plot
df %>%
  group_by(bp_event, Randomization) %>%
  summarise(
    mean_dbp = mean(bp_avgdbp, na.rm = TRUE),
    se_dbp   = sd(bp_avgdbp, na.rm = TRUE) / sqrt(n())
  ) %>%
  ggplot(aes(bp_event, mean_dbp, color = Randomization, group = Randomization)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_dbp - se_dbp,
                    ymax = mean_dbp + se_dbp),
                width = 0.1) +
  labs(
    title = "Mean Diastolic Blood Pressure Over Time by Study Arm",
    x = "Visit",
    y = "Mean Diastolic Blood Pressure (mmHg)",
    Color = "Study Arm"
  ) +
  theme_minimal()
```

## DASH Adherence Score

### Difference-in-Differences Analysis using LMM

```{r dash lmm}
# fit a linear mixed effects model for DASH adherence score by randomization group
lmm_dash <- lmer(
  dash ~ arm * time + (1 | case),
  data = dash_long
)

summary(lmm_dash)

# fit an estimated marginal means model to get group-specific estimates
emm_dash <- emmeans(lmm_dash, ~ arm * time)

contrast(
  emm_dash,
  method = "trt.vs.ctrl",
  ref = "Baseline",
  by = "arm"
)

summary(emm_dash, infer = TRUE)

# fit a difference-in-differences contrast to compare changes in DASH adherence between groups
did_dash <- contrast(
  emm_dash,
  interaction = "revpairwise"
)
summary(did_dash, infer = TRUE)
```

### Data Visualization

```{r dash plot}
dash_long %>%
  mutate(
    time = recode(time,
                  "dash_bl" = "Baseline",
                  "dash_12" = "12 Weeks",
                  "dash_24" = "24 Weeks"),
    time = factor(time, levels = c("Baseline", "12 Weeks", "24 Weeks"))
  ) %>%
  group_by(arm, time) %>%
  summarise(mean_dash = mean(dash, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(time, mean_dash, color = arm, group = arm)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  labs(
    title = "Changes in DASH Adherence Over Time by Study Arm",
    x = "Visit",
    y = "Mean DASH Adherence Score",
    color = "Study Arm"
  ) +
  theme_minimal()
```
